Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-02 talk/dls2014/paper/paper.tex
Some work on the introduction, background

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy, V8, Tracemonkey) started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Another technology that can provide the same guarantees
is transactional memory (TM). \remi{cite our position paper}

There have been several attempts at replacing the GIL with TM. Using
transactions to enclose multiple bytecode instructions, we can get the
very same semantics as the GIL while possibly executing several
transactions in parallel. Furthermore, by exposing these
interpreter-level transactions to the application in the form of
\emph{atomic blocks}, we give dynamic languages a new synchronisation
mechanism that avoids several of the problems of locks as they are
used now.

\remi{cite and extract from (our pos. paper):}
TM systems come in can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints, while STM systems have a lot of overhead.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of hardware features\arigo{"OS-level feature" maybe.
  "Hardware feature" implies it only works on custom chips}
  and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is
  then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

@@

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolation from other
transactions. \emph{Atomicity} means that all these instructions in
the transaction and their effects seem to happen at one, indivisible
point in time. Other transactions never see inconsistent state of a
partially executed transaction which is called \emph{isolation}.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimisation.
\remi{maybe some more explanation of how exactly TM replaces the GIL}

@@

\cfbolz{a pypy introduction needs to go somewhere, a paragraph or so. maybe in the evaluation section}

We implement and evaluate our system for the Python language. For the
actual implementation, we chose the PyPy interpreter because replacing
the GIL there with a TM system is just a matter of adding a new
transformation to the translation process of the interpreter.

Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading} and \emph{multiprocessing}.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython and IronPython are implementations of
this. It requires great care in order to avoid deadlocks, which is why
we follow the TM approach that provides a \emph{direct} replacement
for the GIL. It does not require careful placing of locks in the right
spots. We will compare our work with Jython for evaluation.

@@

\cfbolz{citation again needed for the whole subsection}

It is well known that using locks to synchronise multiple threads is
hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better
parallel programming model for dynamic languages, we want to implement
another, well-known synchronisation mechanism: \emph{atomic blocks}.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. An implementation using a GIL would simply
guarantee that the GIL is not released during the execution of the
atomic block. Using TM, we have the same effect by guaranteeing that
all instructions in an atomic block are executed inside a single
transaction.

\remi{STM, how atomicity \& isolation; reasons for overhead}

@@

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in \ref{fig:performance-nojit}.

@@

The results are shown in \ref{fig:performance-nojit}. We see that the
performance is much less stable. There is certainly more work required
in this area. In general, we see that the group of non-locked
benchmarks certainly scales best. The other three scale barely or not
at all with the number of threads. The slowdown factor from GIL to STM
ranges around \remi{$1-2.4\times$} and we beat GIL performance in half
of the benchmarks.

@@

@@

\bibitem{webjython}
  The Jython Project, \url{www.jython.org}

>>>>

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. Another technology that
can provide the same guarantees is transactional memory
(TM). \remi{cite our position paper}

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}. Using transactions to enclose
multiple bytecode instructions, we can get the very same semantics as
the GIL while possibly executing several transactions in
parallel. Furthermore, by exposing these interpreter-level
transactions to the application in the form of \emph{atomic
blocks}~\cite{tim03,tim05}, we give dynamic languages a new
synchronisation mechanism that avoids several of the problems of locks
as they are used now.

TM systems come in can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of long-existing CPU features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is
  then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

\subsection{Global Interpreter Lock}

The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. Thus, these
instructions are always executed atomically and in complete isolation
from others running in other threads. \emph{Atomicity} means that each
instruction and its effects seem to happen at one, indivisible point
in time. Other instructions never see inconsistent state of a
partially executed instruction (\emph{isolation}).

In addition to these guarantees, instructions are executed in a
sequential consistency model~\cite{lamport79}. This means that
the outcome of any execution of instructions in multiple threads is
equal to \emph{some} sequential execution of them.

@@

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolation from other transactions.
Atomicity and isolation are basic properties of transactions.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. By that, we can again provide a
sequentially consistent model for programming in multiple threads. We
can therefore use TM to directly replace the GIL. Instead of releasing
and acquiring the GIL between bytecode instructions, we commit and
start the transactions our instructions are running in.

@@

We implement and evaluate our system for the Python language. Python
is a dynamic programming language that was designed with GIL semantics
in mind. Its reference implementation, CPython~\cite{cpython}, uses a
GIL to synchronise instructions in multiple threads.
Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading} and \emph{multiprocessing}.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} are implementations of this. It requires
great care in order to avoid deadlocks, which is why we follow the TM
approach that provides a \emph{direct} replacement for the GIL. It
does not require careful placing of locks in the right spots. We will
compare our work with Jython for evaluation.

@@

In Python, since the GIL is not directly exposed to the interpreter,
applications still need to synchronise memory accesses from multiple
threads using locks. Locks can be very hard to get
right~\cite{christopher10,victor11,shan08}.  They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. We think that \emph{atomic
blocks}~\cite{tim03,tim05} provide a better way for synchronisation.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. This is why we think that the introduction
of atomic blocks to Python is a valuable contribution. Since atomicity
is a property of transactions, TM and atomic blocks are a natural fit.

@@

We now take a closer look at how our TM system that we use to replace
the GIL works, what properties it has, and how it is implemented.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in figure \ref{fig:performance-nojit}.

@@

The results are presented in figure \ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. In general, we see that the group of
non-locked benchmarks certainly scales best. The other three scale
barely or not at all with the number of threads. The slowdown factor
from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
performance in half of the benchmarks.

\remi{Reason for bad scaling: acceleration of code that produces
conflicts $-->$ more iterations $-->$ more conflicts. The overhead
doesn't get accelerated by the JIT.}

@@

\bibitem{cfbolz09} Carl Friedrich Bolz, Antonio Cuni, Maciej
  Fijalkowski, and Armin Rigo. 2009. Tracing the meta-level: PyPy's
  tracing JIT compiler.  \emph{In Proceedings of the 4th workshop on the
    Implementation, Compilation, Optimization of Object-Oriented Languages
    and Programming Systems} (ICOOOLPS '09).

\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  A New
  Crankshaft for V8.
  \url{http://blog.chromium.org/2010/12/new-crankshaft-for-v8.html}

\bibitem{ionmonkey} IonMonkey from Mozilla. 2014.
  \url{https://wiki.mozilla.org/IonMonkey/Overview}

\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
  in Parallelising Dynamic Languages. Under review in ICOOOLPS'14.

\bibitem{cpython} CPython. \url{www.python.org}
\bibitem{webjython} The Jython Project, \url{www.jython.org}
\bibitem{ironpython} IronPython. \url{www.ironpython.net}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@
